---
title: "Police Violence Mortality Estimates"
author: "Mike and Frank"
date: "May 8, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract
*Objective.* To estimate mortality for police-involved deaths by race and geography in the United States.

*Methods.* Using data from a crowd-sourced database, we estimate mortality rates for deaths invovling police by race and by census region. Multilevel Bayesian Poisson models adjust for unobserved geographic characteristics and allow for more predictive precision for rare events, such as law enforcement deaths.

*Results.* 

<!-- We identified 1552 police homicides in the 16 states. Positive predictive value and sensitivity of the NVDRS “type of death” variable for police homicides were high (98% and 90%, respectively). Counts from Vital Statistics and Supplementary Homicide Reports were 58% and 48%, respectively, of our study total; gaps varied widely by state. The annual rate of police homicide (0.24/100 000) varied 5-fold by state and 8-fold by race/ethnicity. -->

*Conclusions.* 

*Policy Implications.*


# Introduction

Intro about high-profile cases, need for clear estimates of mortality by group, likelihood of regional heterogeneity in risk, likelihood of msaXregionXrace interaction

# Background

Concise lit review with no more than 35 references

# Data and Methods

We 

# Findings

# Discussion

```{r pressure, echo=FALSE}
plot(pressure)
```

